---
title: "Creating Heatmaps"
author: "Jacob Rozran"
date: "March 5, 2016"
output: html_document
---

# STEP 1: READ IN THE DATA

Get the data - I've just got it downloaded locally. Also - load the required 
libraries for the mapping.

```{r setup}
library(ggplot2)        ## VISUALIZATIONS
library(ggmap)          ## MAPS

## READ IN THE DATA 
conc_deaths <- read.csv("Accidental_Drug_Related_Deaths_January_2012-Sept_2015.csv")
```

# STEP 2: CLEAN THE DATA 

The location data, in the **deathLoc** column, is ugly. It comes in this format:

`r conc_deaths$deathLoc[1]`

To clean it, I first do a substitution for the '\n' into a ';'. R was throwing 
errors when I tried to just split on '\n'. I then split the data on the ';'.

```{r first_split}
## SUBSTITUTE '\n' FOR ';'
loc_data <- gsub(pattern = "\n", replacement = ";", x = conc_deaths$DeathLoc)
## SPLIT ON ';'
loc_data <- strsplit(loc_data, ";")
```

I then add the **town** and **cordinates** to the initial dataset as separate columns 
and blow away the loc_data list and the original **DeathLoc** column (just a 
couple clean up steps).

```{r add_town_and_coord}
## FOR EACH ROW, ADD THE TOWN TO THE TOWN COLUMN 
## AND COORDINATES TO THE COORDINATES COLUMN
for(i in 1:length(loc_data)){
        conc_deaths$town[i] <- loc_data[[i]][1]
        conc_deaths$coordinates[i] <- loc_data[[i]][2]
}

## BLOW AWAY THE NOW UNNEEDED VARIABLES
loc_data <- NULL
conc_deaths$DeathLoc <- NULL
```

Now we need to clean up the coordinates column and split it into a latitude and 
longitude. I then add them to their own columns

```{r split_coord}
## REMOVE THE '(' 
conc_deaths$coordinates <- gsub(pattern = "\\(", replacement = "", 
                          x = conc_deaths$coordinates)
## REMOVE THE ')' 
conc_deaths$coordinates <- gsub(pattern = ")", replacement = "", 
                          x = conc_deaths$coordinates)

## SPLIT ON THE ','
loc_data <- strsplit(conc_deaths$coordinates, split = ", ")

## FOR EACH ROW, ADD THE LATITUDE TO THE LAT COLUMN 
## AND LONGITUDE TO THE LON COLUMN
for(i in 1:length(loc_data)){
        conc_deaths$lat[i] <- loc_data[[i]][1]
        conc_deaths$lon[i] <- loc_data[[i]][2]
}
```

Final clean up step before we map - we need to change the **lat** and **lon** 
columns to numerics - by default they are factors. Also, unnecessarily, I am 
blowing away the unneeded variables. 

```{r as_numeric}
## CHANGE THE VARIABLE TO NUMERICS
conc_deaths$lat <- as.numeric(conc_deaths$lat)
conc_deaths$lon <- as.numeric(conc_deaths$lon)

## GET RID OF THE UNNECESSARY VARIABLES
loc_data <- NULL
conc_deaths$coord <- NULL
```

# MAPPING THE DATA

Finally, we are ready to map the data!

Here we get the map of Conneticut (I do it for 2 levels of zoom), and then map 
it! The higher level of zoom gives you this map:

```{r map1}
## GET THE BASE MAP
con <- get_map(location = "Conneticut", maptype = "roadmap", zoom = 8)

## ADD THE "HEAT" TO THE MAP
map1 <- ggmap(con, extent = "device") + geom_density2d(data = conc_deaths, 
                                                        aes(x = lon, y = lat)) + 
        stat_density2d(data = conc_deaths, aes(fill = ..level.., alpha = ..level..),
                       size = 0.01, geom = "polygon") + 
        scale_fill_gradient(low = "green", high = "red", guide = FALSE) + 
        scale_alpha(range = c(0, 0.3), guide = FALSE)

## DISPLAY IT
map1
```

The next higher zoom gives you this map:

```{r map2}
## GET THE BASE MAP
con1 <- get_map(location = "Conneticut", maptype = "roadmap", zoom = 9)

## ADD THE "HEAT" TO THE MAP
map2 <- ggmap(con1, extent = "device") + geom_density2d(data = conc_deaths, 
                                                        aes(x = lon, y = lat)) + 
        stat_density2d(data = conc_deaths, aes(fill = ..level.., alpha = ..level..),
                       size = 0.01, geom = "polygon") + 
        scale_fill_gradient(low = "green", high = "red", guide = FALSE) + 
        scale_alpha(range = c(0, 0.3), guide = FALSE)

## DISPLAY IT
map2
```